Counterfactual Training: Teaching Models Plausible and Actionable Explanations

### 2025-03-07 paper/preprint/paper.tex
some notes from meeting and updates to appendix

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}

@@

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Counterfactual Training: Teaching Models Plausible and Actionable Explanations},
  pdfauthor={Patrick Altmeyer; Aleksander Buszydlik; Arie van Deursen; Cynthia C. S. Liem},
  pdfkeywords={Counterfactual Explanations, Explainable
AI, Representation Learning},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Arie van
Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel model objective that leverages counterfactual
explanations to increase the explanatory capacity of trained models.
Counterfactual explanations have emerged as a popular tool to explain
predictions made by opaque machine learning models: they inform how
factual inputs would need to change in order for a model to produce some
desired output. Much existing research has focused on generating
counterfactuals that are not only valid but also deemed acceptable with
respect to the stakeholder requirements and plausible with respect to
the underlying data. Recent work has shown that under this premise, the
task of learning plausible explanations---ones that respect the
conditional distribution of the target class---is effectively reassigned
from the model itself to the post-hoc counterfactual explainer. Building
on that work, we propose a novel objective function that employs
counterfactuals (ad-hoc) during the training phase to minimize the
divergence between the learned representations and the explanations.
Through extensive experiments, we demonstrate that our proposed method
facilitates training models that deliver inherently plausible
explanations while maintaining high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Explanations \sep Explainable AI \sep 
Representation Learning

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, coltitle=black, titlerule=0mm, left=2mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, title={Note}, colback=white, rightrule=.15mm, toprule=.15mm, bottomtitle=1mm, opacitybacktitle=0.6, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, coltitle=black, titlerule=0mm, left=2mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colback=white, rightrule=.15mm, toprule=.15mm, bottomtitle=1mm, opacitybacktitle=0.6, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, coltitle=black, titlerule=0mm, left=2mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colback=white, rightrule=.15mm, toprule=.15mm, bottomtitle=1mm, opacitybacktitle=0.6, opacityback=0]

@@

\subsubsection{High-Level Findings}\label{high-level-findings}

@@

@@

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/logs/objective---lambda_energy---generator_type---maxiter/percent_valid.png}}

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/logs/objective---lambda_energy---generator_type---maxiter/acc_val.png}}

\subsubsection{Detailed Findings}\label{detailed-findings}

\paragraph{Linearly Separable}\label{linearly-separable}

\begin{longtable}{cccccc}

\caption{\label{tbl-lin_sep-lambda_energy_exper}Results for Linearly
Separable data by energy penalty.}

\tabularnewline

  \toprule
  \textbf{Decision Threshold} & \textbf{$\lambda_{\text{energy}}$} & \textbf{Maximum Iterations} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
  \endfirsthead
  \toprule
  \textbf{Decision Threshold} & \textbf{$\lambda_{\text{energy}}$} & \textbf{Maximum Iterations} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
  \endhead
  \bottomrule
  \multicolumn{6}{r}{Continuing table below.}\\
  \bottomrule
  \endfoot
  \endlastfoot
  0.75 & 0.1 & 5 & \textit{ECCo} & -15.9 & 16.7 \\
  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{41.9}} & \color{Green}{\textbf{8.13}} \\
  0.75 & 0.1 & 5 & \textit{REVISE} & -64.4 & 3.58 \\
  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{28.9}} & \color{Green}{\textbf{2.28}} \\
  0.75 & 0.1 & 25 & \textit{Generic} & -11.3 & 11.5 \\
  0.75 & 0.1 & 25 & \textit{REVISE} & -55.3 & 4.18 \\
  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{10.9}} & \color{Green}{\textbf{7.47}} \\
  0.75 & 0.1 & 50 & \textit{Generic} & -32.3 & 16.6 \\
  0.75 & 0.1 & 50 & \textit{REVISE} & -44.1 & 4.24 \\
  0.9 & 0.1 & 5 & \textit{ECCo} & -13.7 & 13.7 \\
  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{22.3}} & \color{Green}{\textbf{4.87}} \\
  0.9 & 0.1 & 5 & \textit{REVISE} & -27.8 & 6.26 \\
  0.9 & 0.1 & 25 & \textit{ECCo} & 8.84 & 7.13 \\
  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{30.7}} & \color{Green}{\textbf{5.77}} \\
  0.9 & 0.1 & 25 & \textit{REVISE} & -72.7 & 34.9 \\
  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{17.1}} & \color{Green}{\textbf{7.98}} \\
  0.9 & 0.1 & 50 & \textit{Generic} & 3.82 & 8.86 \\
  0.9 & 0.1 & 50 & \textit{REVISE} & -45.8 & 2.79 \\
  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{35.7}} & \color{Green}{\textbf{7.56}} \\
  0.95 & 0.1 & 5 & \textit{Generic} & -13.3 & 5.73 \\
  0.95 & 0.1 & 5 & \textit{REVISE} & -54.9 & 5.43 \\
  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{37}} & \color{Green}{\textbf{5.31}} \\
  0.95 & 0.1 & 25 & \textit{Generic} & 29 & 8.6 \\
  0.95 & 0.1 & 25 & \textit{REVISE} & -40.5 & 41 \\
  0.95 & 0.1 & 50 & \textit{ECCo} & 13.2 & 4.89 \\
  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{35.5}} & \color{Green}{\textbf{4.27}} \\
  0.95 & 0.1 & 50 & \textit{REVISE} & -60.4 & 45.8 \\
  \color{Red}{\textbf{0.75}} & \color{Red}{\textbf{0.5}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{ECCo}} & \color{Red}{\textbf{-0.94}} & \color{Red}{\textbf{17.6}} \\
  0.75 & 0.5 & 5 & \textit{Generic} & -11.7 & 17.4 \\
  0.75 & 0.5 & 5 & \textit{REVISE} & -37.7 & 3.92 \\
  0.75 & 0.5 & 25 & \textit{ECCo} & 14.7 & 6.32 \\
  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{32.3}} & \color{Green}{\textbf{3.49}} \\
  0.75 & 0.5 & 25 & \textit{REVISE} & -39.9 & 7.21 \\
  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{28.3}} & \color{Green}{\textbf{5.05}} \\
  0.75 & 0.5 & 50 & \textit{Generic} & -9.72 & 9.19 \\
  0.75 & 0.5 & 50 & \textit{REVISE} & -30 & 5.22 \\
  0.9 & 0.5 & 5 & \textit{ECCo} & -7.38 & 13.6 \\
  \color{Red}{\textbf{0.9}} & \color{Red}{\textbf{0.5}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{Generic}} & \color{Red}{\textbf{-0.0952}} & \color{Red}{\textbf{17.1}} \\
  0.9 & 0.5 & 5 & \textit{REVISE} & -50.6 & 4.77 \\
  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{39.1}} & \color{Green}{\textbf{8.3}} \\
  0.9 & 0.5 & 25 & \textit{Generic} & 38.6 & 6.03 \\
  0.9 & 0.5 & 25 & \textit{REVISE} & -56.9 & 4.28 \\
  0.9 & 0.5 & 50 & \textit{ECCo} & -6.53 & 10.9 \\
  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{13.6}} & \color{Green}{\textbf{12.8}} \\
  0.9 & 0.5 & 50 & \textit{REVISE} & -24.1 & 12.5 \\
  0.95 & 0.5 & 5 & \textit{ECCo} & 0.596 & 1.11 \\
  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{4.57}} & \color{Green}{\textbf{5.88}} \\
  0.95 & 0.5 & 5 & \textit{REVISE} & -34.3 & 6.09 \\
  0.95 & 0.5 & 25 & \textit{ECCo} & 16.5 & 3.53 \\
  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{19.1}} & \color{Green}{\textbf{4.43}} \\
  0.95 & 0.5 & 25 & \textit{REVISE} & -37.2 & 39.2 \\
  0.95 & 0.5 & 50 & \textit{ECCo} & -31 & 14.2 \\
  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{24.8}} & \color{Green}{\textbf{4.78}} \\
  0.95 & 0.5 & 50 & \textit{REVISE} & -1.28 & 2.7 \\
  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{37}} & \color{Green}{\textbf{9.21}} \\
  0.75 & 5 & 5 & \textit{Generic} & 0.977 & 21.1 \\
  0.75 & 5 & 5 & \textit{REVISE} & -55.2 & 4.75 \\
  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{21.8}} & \color{Green}{\textbf{5.97}} \\
  0.75 & 5 & 25 & \textit{Generic} & 17.8 & 5.33 \\
  0.75 & 5 & 25 & \textit{REVISE} & -68.1 & 8.69 \\
  0.75 & 5 & 50 & \textit{ECCo} & 0.864 & 9.51 \\
  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{8.56}} & \color{Green}{\textbf{13.6}} \\
  0.75 & 5 & 50 & \textit{REVISE} & -71.8 & 3.18 \\
  0.9 & 5 & 5 & \textit{ECCo} & -3.59 & 20.3 \\
  \color{Red}{\textbf{0.9}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{Generic}} & \color{Red}{\textbf{-1.18}} & \color{Red}{\textbf{16.7}} \\
  0.9 & 5 & 5 & \textit{REVISE} & -58.6 & 3.55 \\
  0.9 & 5 & 25 & \textit{ECCo} & -1.73 & 9.07 \\

>>>>

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}

@@

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Counterfactual Training: Teaching Models Plausible and Actionable Explanations},
  pdfauthor={Patrick Altmeyer; Aleksander Buszydlik; Arie van Deursen; Cynthia C. S. Liem},
  pdfkeywords={Counterfactual Training, Counterfactual
Explanations, Algorithmic Recourse, Explainable AI, Representation
Learning},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Arie van
Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel training regime called Counterfactual Training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to
stakeholder requirements. Much existing research has therefore focused
on developing post-hoc methods to generate counterfactuals that meet
these desiderata. In this work, we instead hold models directly
accountable for this desired end goal: Counterfactual Training employs
counterfactuals ad-hoc during the training phase to minimize the
divergence between learned representations and plausible, actionable
explanations. We demonstrate empirically and theoretically that our
proposed method facilitates training models that deliver inherently
desirable explanations while maintaining high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Training \sep Counterfactual
Explanations \sep Algorithmic Recourse \sep Explainable AI \sep 
Representation Learning

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, left=2mm, coltitle=black, toptitle=1mm, breakable, bottomtitle=1mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, title={Note}, titlerule=0mm, arc=.35mm, toprule=.15mm, rightrule=.15mm, opacityback=0, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, left=2mm, coltitle=black, toptitle=1mm, breakable, bottomtitle=1mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, titlerule=0mm, arc=.35mm, toprule=.15mm, rightrule=.15mm, opacityback=0, bottomrule=.15mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, left=2mm, coltitle=black, toptitle=1mm, breakable, bottomtitle=1mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, titlerule=0mm, arc=.35mm, toprule=.15mm, rightrule=.15mm, opacityback=0, bottomrule=.15mm]

@@

\paragraph{Plausibility}\label{plausibility-1}

The results with respect to the plausibility measure are shown in
Figure~\ref{fig-grid-gen_params-plaus-circles} to
Figure~\ref{fig-grid-gen_params-plaus-over}.

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/circles/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}

\caption{\label{fig-grid-gen_params-plaus-circles}Average outcomes for
the plausibility measure across hyperparameters. Data: Circles.}

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/lin_sep/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}

\caption{\label{fig-grid-gen_params-plaus-lin_sep}Average outcomes for
the plausibility measure across hyperparameters. Data: Linearly
Separable.}

@@

\caption{\label{fig-grid-gen_params-plaus-moons}Average outcomes for the
plausibility measure across hyperparameters. Data: Moons.}

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/over/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}

\caption{\label{fig-grid-gen_params-plaus-over}Average outcomes for the
plausibility measure across hyperparameters. Data: Overlapping.}

\paragraph{Cost}\label{cost}

The results with respect to the cost measure are shown in
Figure~\ref{fig-grid-gen_params-cost-circles} to
Figure~\ref{fig-grid-gen_params-cost-over}.

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/circles/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}

\caption{\label{fig-grid-gen_params-cost-circles}Average outcomes for
the cost measure across hyperparameters. Data: Circles.}

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/lin_sep/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}

\caption{\label{fig-grid-gen_params-cost-lin_sep}Average outcomes for
the cost measure across hyperparameters. Data: Linearly Separable.}

@@

\caption{\label{fig-grid-gen_params-cost-moons}Average outcomes for the
cost measure across hyperparameters. Data: Moons.}

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/over/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}

\caption{\label{fig-grid-gen_params-cost-over}Average outcomes for the
cost measure across hyperparameters. Data: Overlapping.}

\subsubsection{Penalty Strengths}\label{penalty-strengths}

The hyperparameter grid with varying penalty strengths during training
is shown in Note~\ref{nte-pen-final-run-train}. The corresponding
evaluation grid used for these experiments is shown in
Note~\ref{nte-pen-final-run-eval}.

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, left=2mm, coltitle=black, toptitle=1mm, breakable, bottomtitle=1mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, titlerule=0mm, arc=.35mm, toprule=.15mm, rightrule=.15mm, opacityback=0, bottomrule=.15mm]

\quartocalloutnte{nte-pen-final-run-train}

\begin{itemize}
\tightlist
\item
  Generator: \texttt{ecco,\ generic,\ revise}
\item
  Model: \texttt{mlp}
\item
  Training Parameters:

  \begin{itemize}
  \tightlist
  \item
    \(\lambda_{\text{adv}}\): \texttt{0.1,\ 0.25,\ 1.0}
  \item
    \(\lambda_{\text{div}}\): \texttt{0.01,\ 0.1,\ 1.0}
  \item
    \(\lambda_{\text{reg}}\): \texttt{0.0,\ 0.01,\ 0.1}
  \item
    Objective: \texttt{full,\ vanilla}
  \end{itemize}
\end{itemize}

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, left=2mm, coltitle=black, toptitle=1mm, breakable, bottomtitle=1mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, titlerule=0mm, arc=.35mm, toprule=.15mm, rightrule=.15mm, opacityback=0, bottomrule=.15mm]

\quartocalloutnte{nte-pen-final-run-eval}

\paragraph{Plausibility}\label{plausibility-2}

The results with respect to the plausibility measure are shown in
Figure~\ref{fig-grid-pen-plaus-circles} to
Figure~\ref{fig-grid-pen-plaus-over}.

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/circles/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}

\caption{\label{fig-grid-pen-plaus-circles}Average outcomes for the
plausibility measure across hyperparameters. Data: Circles.}

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/lin_sep/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}

\caption{\label{fig-grid-pen-plaus-lin_sep}Average outcomes for the
plausibility measure across hyperparameters. Data: Linearly Separable.}

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/moons/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}

\caption{\label{fig-grid-pen-plaus-moons}Average outcomes for the
plausibility measure across hyperparameters. Data: Moons.}

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/over/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}

\caption{\label{fig-grid-pen-plaus-over}Average outcomes for the
plausibility measure across hyperparameters. Data: Overlapping.}

\paragraph{Cost}\label{cost-1}

The results with respect to the cost measure are shown in
Figure~\ref{fig-grid-pen-cost-circles} to
Figure~\ref{fig-grid-pen-cost-over}.

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/circles/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}

\caption{\label{fig-grid-pen-cost-circles}Average outcomes for the cost
measure across hyperparameters. Data: Circles.}

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/lin_sep/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}

\caption{\label{fig-grid-pen-cost-lin_sep}Average outcomes for the cost
measure across hyperparameters. Data: Linearly Separable.}

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/moons/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}

\caption{\label{fig-grid-pen-cost-moons}Average outcomes for the cost
measure across hyperparameters. Data: Moons.}

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/over/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}

\caption{\label{fig-grid-pen-cost-over}Average outcomes for the cost
measure across hyperparameters. Data: Overlapping.}
